Light inextensible strings under tension in the Schwarzschild geometry

### 2019-06-07 string_grg.tex
initial draft using the style for <General Relativity and Gravitation>

\documentclass[prb,preprint]{revtex4-1} 
% The line above defines the type of LaTeX document.
% Note that AJP uses the same style as Phys. Rev. B (prb).

% The % character begins a comment, which continues to the end of the line.

\usepackage{amsmath}  % needed for \tfrac, \bmatrix, etc.
\usepackage{amsfonts} % needed for bold Greek, Fraktur, and blackboard bold
\usepackage{graphicx} % needed for figures

\begin{document}

% Be sure to use the \title, \author, \affiliation, and \abstract macros
% to format your title page.  Don't use lower-level macros to  manually
% adjust the fonts and centering.

\title{Light inextensible strings under tension in the Schwarzschild geometry}
% In a long title you can use \\ to force a line break at a certain location.

\author{Robin K. S. Hankin}
\email{hankin.robin@gmail.com} % optional
\altaffiliation[permanent address: ]{55 Wellesley Street East,
Auckland, New Zealand}
% If there were a second author at the same address, we would put another 
% \author{} statement here.  Don't combine multiple authors in a single
% \author statement.
\affiliation{School of Engineering, Computer and Mathematical Sciences, Auckland University of Technology}
% Please provide a full mailing address here.

% See the REVTeX documentation for more examples of author and affiliation lists.

\date{\today}

\begin{abstract}
Here I analyze the behaviour of stationary, light inextensible strings
under tension in the vicinity of a nonrotating black hole.  Such
strings minimize their proper length, given by integrating the spatial
component of the Schwarzschild metric along the string.  The path is
calculated by numerical solution to the Euler-Lagrange equations.  If
the string is entirely outside the event horizon, its closest approach
to the singularity is tangential; if the distance of closest
approach is less than about $1.0761\cdot 2M$, the
string self-intersects.  If the string attains $r=2M$, then it
will in general be tangential to the event horizon.  Light taut
strings interpret the concept ``straight line'' and their behaviour
suggests a possible way of manipulating a black hole.
\end{abstract}
% AJP requires an abstract for all regular article submissions.
% Abstracts are optional for submissions to the "Notes and Discussions" section.

\maketitle % title page is now complete

@@

\label{intro}
Light inextensible strings are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
expressions for the strength of gravity at the event horizon of a
black hole gives $L\lesssim TM/\left(\mu c^2\right)$.  For the string
to behave as though light on lengthscales comparable with the
Schwarzschild radius would require $T\gtrsim\mu c^2$.  Exceeding this
criterion would violate the dominant energy condition (and in this
context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
string as light and inextensible is an interesting and informative
limit: its analysis interprets the concept ``straight line'' and
suggests a possible way of manipulating a black hole.  Peculiar
counterintuitive behaviour such as inward-directed centrifugal force
in the vicinity of a black hole is well
known~\citep{abramowicz1992,abramowicz1993}, and the analysis of taut
strings furnishes another everyday window into general relativity.
Another possibility, presented by \citet{damour1978} and discussed by
\citet{thorne1986}, is to pass electric current through such taut
strings, an option not pursued here.

@@

%%Vancouver style references.
\bibliographystyle{model3-num-names}
\bibliography{stringrefs}

@@

Calculations are performed using the R programming language~\citep{rcore2018};
source code for the figures is available at
\\
\\
\url{https://github.com/RobinHankin/string.git}

@@

\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{amssymb}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
%\usepackage{latexsym}
% etc.
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Light inextensible strings under tension in the Schwarzschild geometry}%\thanks{Grants or other notes
%about the article that should go on the front page should be
%placed here. General acknowledgments should be placed at the end of the article.}

%\subtitle{Do you have a subtitle?\\ If so, write it here}

%\titlerunning{Short form of title}        % if too long for running head

\author{Robin K. S. Hankin}

%\authorrunning{Short form of author list} % if too long for running head

\institute{Robin K. S. Hankin \at
  Auckland University of Technology\\
  Wellesley Street\\
  Auckland, New Zealand\\
              Tel.: +64(9)5219999\\
              \email{hankin.robin@gmail.com}           %  \\
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}

  Here I analyze the behaviour of stationary, light inextensible strings
under tension in the vicinity of a nonrotating black hole.  Such
strings minimize their proper length, given by integrating the spatial
component of the Schwarzschild metric along the string.  The path is
calculated by numerical solution to the Euler-Lagrange equations.  If
the string is entirely outside the event horizon, its closest approach
to the singularity is tangential; if the distance of closest
approach is less than about $1.0761\cdot 2M$, the
string self-intersects.  If the string attains $r=2M$, then it
will in general be tangential to the event horizon.  Light taut
strings interpret the concept ``straight line'' and their behaviour
suggests a possible way of manipulating a black hole.
\keywords{Black holes\and inextensible string\and minimal-length path}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}

@@

\label{intro}
Light inextensible strings are a stalwart feature of elementary
physics.  A light string is not subject to gravity, and being under
tension means that its configuration is an extremum of proper length.
Of course, no real string is perfectly weightless or inextensible.
With tension~$T$, gravitational acceleration $g$ and (proper) mass per
(proper) unit length~$\mu$, then a string will behave as though light
on lengthscales $L\lesssim T/\left(\mu g\right)$.  Using Newtonian
expressions for the strength of gravity at the event horizon of a
black hole with mass $M$ gives $L\lesssim TM/\left(\mu c^4\right)$.
For the string to behave as though light on lengthscales comparable
with the Schwarzschild radius would require $T\gtrsim\mu c^2$.
Exceeding this criterion would violate the dominant energy condition
(and in this context we note that transverse waves would propagate at
speed~$\sqrt{T/\mu}\sim c$).  Nevertheless, the idealization of a
string as light and inextensible is an interesting and informative
limit: its analysis interprets the concept ``straight line'' and
suggests a possible way of manipulating a black hole.  Peculiar
counterintuitive behaviour such as inward-directed centrifugal force
in the vicinity of a black hole is well
known~\cite{abramowicz1992,abramowicz1993}, and the analysis of taut
strings furnishes another everyday window into general relativity.
Another possibility, presented by discussed by
Thorne~\cite{thorne1986}, is to pass electric current through such
taut strings, an option not pursued here.

@@

%%Vancouver style references.

@@

Calculations are performed using the R programming language~\cite{rcore2018};
source code for the figures is available at
\\
\\
{\tt https://github.com/RobinHankin/string.git}

@@

 \section*{Conflict of interest}
 The author declares that he has no conflict of interest.

% BibTeX users please use one of
%\bibliographystyle{spbasic}      % basic style, author-year citations
\bibliographystyle{spmpsci}      % mathematics and physical sciences
%\bibliographystyle{spphys}       % APS-like style for physics
%\bibliography{}   % name your BibTeX data base

\bibliography{stringrefs}

\end{document} 
% end of file template.tex
